Reasoning Graphically about Quantum Computation

### 2008-02-26 papers/reasoning-graphically/paper.tex
updated paper and .gvlock for svn:ignore
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1373 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

The representation of Compact Closed Categories as Graphs, discuseed
in the previous section, is too restrictive for automated reasoning
about quantum computation. In particular, pictures such as the
following are frequently needed:

\begin{center}
  \scalebox{1.0}{\includegraphics{images/spider_lhs}}
\end{center}

\noindent Such pictures represent an infinite family of concrete
graphs. In this section we describe an extension of the representation
for Compact Closed Categories that allows a graph-based representation
that captures an infinite family of graphs. Notably, this allows us to
represent laws about quantum computation, such as
Spider-Law~\ref{spider-law-fig}. Furthermore, this representation is
still compositional in the sense that we can compose graphs. The
extension, which we call \ref{graph patterns}, adds two extra kinds of
nodes to the graph. Namely, {\em banged boxes} which informally
represent an arbitrary number of copies of a node (where the node is
defined as some graph), and {\em variable nodes} which informally
represent some, as yet unkown, node in a graph.

@@

The set of concrete graphs for a graph $G$ with !-boxes, is denoted by
$\binterp{G}$.

\subsubsection{Making !-Box Graphs}

@@

The purpose of our graph patterns is to express and formalise, in a
declarative manner, the intuitive idea of graph matching for graphs
that represent Compact Closed Categories. In this section we combine
the two extensions to the concrete graph representation, namely
variable nodes and !-boxes, to provide a matching relation between
graph patterns and concrete graphs. We then introduce a matching
relation directly between graph patterns. The former relation gives a
semantics for our system and ensures soundness, while the latter
supports deirved equations and thus provides the theoretical basis for
our formulation of rewriting.

@@

We give a semantics for a pattern graph $G$ in terms of a set of
concrete graphs, denoted by $\minterp{G}$, and say that members of
this set are \emph{matched} by $G$. Formally, the set of matches are
defined as follows:

$\minterp{G} = \sinterp{\vinterp{\binterp{G}}}$

\noindent where $\sinterp{G}$ represents the set of graphs that
strictly contain the graph $G$. Thus, a concrete graph $g$ is matched
by a graph pattern $G$, written $g \in G$, when an instance of the
pattern is a strict subgraph of the matched graph.

@@

The definition of one pattern matching another one is simply that it
is more general. Formally, a graph pattern, $G$, matches another graph
pattern, $H$, when $\minterp{G} \supseteq \minterp{H}$. This provides
the specification for a matching algorithm which we provide an
implementation of in~\S\ref{sec:rewriting}.

Our representation has been carefully designed so that this concept
corresponds to what we call a \emph{respectful embedding} of $G$ into
$H$.

>>>>

The representation of Compact Closed Categories as Graphs, discuseed
in the previous section, is too restrictive for automated reasoning
about quantum computation. In particular, pictures such the Spider
Law, in Figure~\ref{spider-law-fig} are frequently needed. Such
pictures represent an infinite family of concrete graphs. In this
section we describe an extension of the representation for Compact
Closed Categories that allows a graph-based representation that
captures an infinite family of graphs. Notably, this allows us to
represent laws about quantum computation, such as the Spider
Law. Furthermore, this representation is still compositional in the
sense that we can plug graphs together. The extension, which we call
\ref{graph patterns}, adds two extra kinds of nodes to the
graph. Namely, {\em banged boxes} which informally represent an
arbitrary number of copies of a node (where the node itself can
contain a graph of the appropriate type), and {\em variable nodes}
which informally represent some, as yet unkown, node in a graph.

%\begin{center}
%  \scalebox{1.0}{\includegraphics{images/spider_lhs}}
%\end{center}

@@

The key feature of rules like the Spider Law is that an arbirary
number of nodes can be matched. To support this we introduce an
operation, !-boxing, on graph representations. We draw !-boxes around
a set of chosen nodes and intuitively think of the resulting !-boxed
graph as representing a set of graphs with an arbirary number of
copies of the !-boxed nodes, where any connections to nodes outside
the !-box are also copied. The !-boxes can also be instantiated with
zero copies errasing all edges from the !-box. See
Figure~\ref{bang-box-ex1-fig} for a simple example.

A graph with !-boxes is simply a graph paired with a set of !-box node
sets, which define the !-boxes. The !-box node sets must either be
disjoint or strict subsets of each other. This is needed to avoid
confusion with overlaping boxes.

\subsubsection{Semantics.}

We give a formal semantics to !-boxing a representation of graphs by
defining the a !-box graph in terms of a set of graphs in the
underlying representation. We denote this interpretation of a graph
$G$ by $\binterp{G}$ and say that mebemrs of $\binterp{G}$ are
instances of $G$. The definition of $\binterp{G}$ involves the
following three operations\footnote{These correspond closely to the
  linear logic rules for weakening, contraction and deriliction.} on a
!-box and a graph that contains it:

\begin{definition}
  \emph{copy} copies a box in a graph to produce a new graph with two
  copies of the boxed subgraph, with a second box around the newly
  copied nodes. Any edges incident on both a node in the !-box and
  outside it get copied so that there the old edge remains and the new
  edge goes from the outer node to the new copy of the internal node.
\end{definition}

\begin{definition}
  \emph{drop} simply removes the box, but leaves its contents in the
  graph.
\end{definition}

\begin{definition}
  \emph{kill} removes from the graph all nodes in the !-box set as
  well as any incident edges to nodes in the box.
\end{definition}

\noindent The semantics of a !-boxed graph is then defined as the
members without any !-boxes of the set closed by these operation.  For
the purposes of this paper, the underlying graph representation of
!-boxed graphs is variable-node graphs. The set of variable-node
graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$.

\subsubsection{}

@@

The purpose of graph patterns is to formally express, in a declarative
manner, the intuitive idea of graph matching commonly used in for
Quantum computation. To do this, we combine the two extensions to the
concrete graph representation, namely variable nodes and !-boxes. This
provides a matching relation between graph patterns and concrete
graphs as well as directly between graph patterns. The former relation
gives a semantics for our system and ensures soundness, while the
latter supports deirved equations and thus provides the theoretical
basis for our formulation of rewriting.

@@

The semantics for a graph pattern $G$ in terms of a set of concrete
graphs, is denoted by $\minterp{G}$. We say that members of this set
are \emph{matched} by $G$. Formally, the set of matches are defined as
follows:

$\minterp{G} = (\vinterp{G'} | G' \in \binterp{G})$

@@

The definition of one pattern matching another one is simply that it
is more general. Formally, a graph pattern, $G$, matches another graph
pattern, $H$, when $\minterp{G} \supseteq \minterp{H}$. This provides
the specification for a matching algorithm which we implement of
in~\S\ref{sec:rewriting}. To make such a proof easier, we extend
embedding and plugging of variable-node graphs to graph patterns.

Our representation has been carefully designed so that this concept
corresponds to what we call a \emph{open embedding} of $G$ into
$H$.
